# Results

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(lubridate)
library(scales)
library(here)
library(viridis)
library(zoo)

df <- read_csv(here("data", "clean", "crypto_final_clean.csv"), show_col_types = FALSE)
```

## Overview

This chapter explores how price movements, sentiment, and search attention evolve over time for Bitcoin and Ethereum. The analysis is descriptive and focuses on answering three questions:

1.  How do prices, sentiment, and public attention change over time?
2.  How closely do sentiment and search interest track market volatility?
3.  Do Bitcoin and Ethereum respond differently to behavioral signals?

The chapter begins by establishing baseline market behavior through price and volatility trends. It then examines sentiment as a behavioral indicator, exploring how online attitudes evolve alongside market movements and whether sentiment provides any leading signal for returns. The same structure is applied to search interest. Finally, the chapter compares how these behavioral signals relate to each asset's return distribution.

------------------------------------------------------------------------

## How Did Bitcoin and Ethereum Prices Move This Year?

Before examining behavioral signals, we establish how prices evolved. This baseline helps interpret the sentiment and search patterns that follow.

```{r}
#| label: fig-price-trends
#| fig-cap: "Both coins follow similar seasonal patterns, declining early in the year, rallying through summer, and pulling back in fall."
#| message: false
#| warning: false

ggplot(df, aes(x = date, y = price, color = coin)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_manual(
    values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")
  ) +
  facet_wrap(~ coin, ncol = 1, scales = "free_y") +
  labs(
    x = NULL,
    y = "Price (USD)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Bitcoin and Ethereum move through the year in similar waves. Both decline early, recover strongly in summer, and pull back again in fall. Bitcoin begins near \$100,000, dips to the low \$80,000s in spring, then rallies above \$120,000 before softening. Ethereum follows the same pattern with larger percentage swings, falling from around \$4,000 to below \$2,000 early in the year before nearly doubling off its lows.

This parallel movement reflects how closely both markets respond to the same macroeconomic forces and investor sentiment. Neither coin stays stable for long.

## When Were Markets Most Volatile?

Price direction tells only part of the story. Volatility captures how much prices swing day to day. Higher volatility means larger and more frequent price changes.

```{r}
#| label: fig-volatility
#| fig-cap: "Volatility spikes in spring and fall for both coins, with Ethereum consistently showing larger swings than Bitcoin."
#| message: false
#| warning: false

df_vol <- df |>
  mutate(daily_return = as.numeric(daily_return)) |>
  group_by(coin) |>
  mutate(
    abs_return = abs(daily_return),
    roll_vol = rollapply(abs_return, width = 14, FUN = mean,
                         fill = NA, align = "right")
  )

ggplot(df_vol, aes(x = date, y = roll_vol, color = coin)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ coin, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = label_percent()) +
  labs(
    x = NULL,
    y = "Two-Week Rolling Volatility"
  ) +
  scale_color_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Both assets move through clear periods of calm and turbulence. Volatility rises in spring and again in fall, matching moments when markets reacted to shifting economic news. Ethereum has consistently higher volatility throughout the year. Bitcoin is steadier, with smaller and less frequent spikes. The midsummer volatility increase for both coins appears shortly after their price rebounds, showing that rallies can bring uncertainty alongside optimism.

------------------------------------------------------------------------

## Does Online Sentiment Track Market Movements?

Sentiment captures the tone of online discussions, positive values indicate optimism, negative values indicate concern, and values near zero reflect neutral attitudes. If sentiment moves with prices, it suggests that online mood reflects market conditions.

```{r}
#| label: fig-sentiment-time
#| fig-cap: "Sentiment rises during the summer rally and falls during the spring and fall declines, following price cycles rather than leading them."
#| message: false
#| warning: false

ggplot(df, aes(x = date, y = sentiment_score, color = coin)) +
  geom_line(alpha = 0.5, linewidth = 0.7) +
  geom_smooth(se = FALSE, linewidth = 1.2, span = 0.15) +
  facet_wrap(~ coin, ncol = 1, scales = "free_y") +
  labs(
    x = NULL,
    y = "Sentiment Score"
  ) +
  scale_color_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Sentiment for both coins moves within a fairly narrow band, with frequent short spikes. The broader trend rises through summer and dips in fall, aligning with price cycles. Ethereum shows slightly sharper swings, suggesting its online discussions respond more strongly to news. Overall, sentiment follows price movements rather than leading them, consistent with investors reacting to events as they unfold.

## Are Big Price Swings Accompanied by Extreme Sentiment?

If sentiment relates to market activity, we would expect larger price moves to coincide with more emotional reactions. The violin plots below group days by the size of their price move and show how sentiment is distributed within each group.

```{r}
#| label: fig-sentiment-volatility
#| fig-cap: "On calm days, sentiment clusters near neutral. On volatile days, sentiment spreads wider—some investors turn very optimistic while others turn very concerned."
#| message: false
#| warning: false

df_binned <- df |>
  mutate(
    return_bin = cut(
      abs(daily_return),
      breaks = c(0, 0.01, 0.03, 0.06, Inf),
      labels = c("Very Small\n(<1%)", "Small\n(1-3%)", "Medium\n(3-6%)", "Large\n(>6%)"),
      include.lowest = TRUE
    )
  ) |>
  filter(!is.na(return_bin))

ggplot(df_binned, aes(x = return_bin, y = sentiment_score, fill = coin)) +
  geom_violin(alpha = 0.7, trim = TRUE, color = "gray40", linewidth = 0.3) +
  facet_wrap(~ coin) +
  scale_fill_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = "Size of Daily Price Move",
    y = "Sentiment Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 10)
  )
```

For both coins, sentiment stays close to neutral when price moves are very small or small. As moves grow into the medium and large categories, sentiment spreads wider in both directions. Larger swings produce both more positive and more negative reactions rather than shifting sentiment uniformly one way. Ethereum shows a stronger increase in variation than Bitcoin, consistent with its more reactive market behavior.

## Can Yesterday's Sentiment Predict Today's Returns?

The plots above show that sentiment moves alongside prices. But does sentiment provide any signal about what comes next? The scatterplots below compare yesterday's sentiment with today's return.

```{r}
#| label: fig-sentiment-lag
#| fig-cap: "More positive sentiment tends to precede slightly higher returns the next day, though the relationship is weak. The effect is stronger for Ethereum."
#| message: false
#| warning: false

df_lag <- df |>
  group_by(coin) |>
  mutate(
    sentiment_lag1 = lag(sentiment_score, 1),
    search_lag1 = lag(search_interest, 1)
  )

ggplot(df_lag, aes(x = sentiment_lag1, y = daily_return, color = coin)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1.2, method = "loess") +
  facet_wrap(~ coin) +
  scale_y_continuous(labels = label_percent()) +
  scale_color_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = "Sentiment (Previous Day)",
    y = "Daily Return"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Most points cluster near zero on both axes, typical sentiment changes do not correspond to large next-day moves. However, the smooth trends curve upward for each asset, suggesting that positive sentiment is often followed by slightly higher returns. Ethereum shows a stronger upward curve and wider spread of returns. The effect is modest and not predictive in a practical sense, but sentiment does appear more connected to Ethereum's short-term movements than Bitcoin's.

```{r}
#| label: fig-sentiment-bars
#| fig-cap: "Grouping days by sentiment level confirms the pattern: high sentiment precedes positive average returns for Ethereum, while Bitcoin shows little difference across groups."
#| message: false
#| warning: false

df_bins <- df_lag |>
  mutate(
    sent_bin = cut(
      sentiment_lag1,
      breaks = quantile(sentiment_lag1, probs = c(0, .33, .66, 1), na.rm = TRUE),
      labels = c("Low", "Medium", "High"),
      include.lowest = TRUE
    )
  )

sent_summary <- df_bins |>
  filter(!is.na(sent_bin)) |>
  group_by(coin, sent_bin) |>
  summarize(mean_return = mean(daily_return, na.rm = TRUE), .groups = "drop")

ggplot(sent_summary, aes(x = sent_bin, y = mean_return, fill = coin)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "gray50") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = "Previous-Day Sentiment Level",
    y = "Mean Next-Day Return",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

For Bitcoin, returns hover near zero regardless of previous-day sentiment. Ethereum shows a clearer gradient: low sentiment precedes small losses on average, while high sentiment is followed by noticeably positive returns. This reinforces the finding that Ethereum responds more to shifts in sentiment than Bitcoin does.

------------------------------------------------------------------------

## Does Search Interest Tell a Similar Story?

Search interest captures how often people seek information about each coin, a proxy for public attention. The scale runs from 0 to 100, where 100 represents peak search activity during the year.

```{r}
#| label: fig-search-time
#| fig-cap: "Search interest spikes during periods of high volatility and major news, with Bitcoin showing steadier attention and Ethereum experiencing sharper bursts."
#| message: false
#| warning: false

ggplot(df, aes(x = date, y = search_interest, color = coin)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ coin, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = NULL,
    y = "Search Interest (0-100)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Search interest moves in noticeable waves across the year. Peaks appear during periods of fast price changes or major news. Bitcoin shows strong attention early in the year and again in midsummer and late fall. Ethereum follows similar cycles with sharper bursts. Overall, search activity rises when markets are active or uncertain.

## Is Search Interest a Weaker Signal Than Sentiment?

To test whether search interest provides a leading signal, we repeat the lagged analysis.

```{r}
#| label: fig-search-lag
#| fig-cap: "Unlike sentiment, search interest shows almost no relationship with next-day returns. The trend lines are nearly flat for both coins."
#| message: false
#| warning: false

ggplot(df_lag, aes(x = search_lag1, y = daily_return, color = coin)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1.2, method = "loess") +
  facet_wrap(~ coin) +
  scale_y_continuous(labels = label_percent()) +
  scale_color_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = "Search Interest (Previous Day)",
    y = "Daily Return"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Most points lie close to zero on both axes. The trend lines are nearly flat for Bitcoin and only gently sloped for Ethereum. Higher search interest does not consistently precede stronger price movements for either coin. Compared with sentiment, search interest provides an even weaker signal.

```{r}
#| label: fig-search-bars
#| fig-cap: "Average next-day returns show no consistent pattern across search levels, confirming that search interest is a poor leading indicator."
#| message: false
#| warning: false

df_bins2 <- df_lag |>
  mutate(
    search_bin = cut(
      search_lag1,
      breaks = quantile(search_lag1, probs = c(0, .33, .66, 1), na.rm = TRUE),
      labels = c("Low", "Medium", "High"),
      include.lowest = TRUE
    )
  )

search_summary <- df_bins2 |>
  filter(!is.na(search_bin)) |>
  group_by(coin, search_bin) |>
  summarize(mean_return = mean(daily_return, na.rm = TRUE), .groups = "drop")

ggplot(search_summary, aes(x = search_bin, y = mean_return, fill = coin)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "gray50") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = "Previous-Day Search Level",
    y = "Mean Next-Day Return",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

Bitcoin returns remain near zero across all search levels. Ethereum shows a small rise after medium search interest that disappears at high levels. Spikes in attention do not consistently translate into stronger price reactions. Search interest provides only weak and inconsistent signals compared with sentiment.

------------------------------------------------------------------------

## Why Does Ethereum React More Strongly Than Bitcoin?

Throughout this analysis, Ethereum has shown larger responses to behavioral signals than Bitcoin. The return distributions below help explain why.

```{r}
#| label: fig-return-dist
#| fig-cap: "Ethereum's return distribution is wider with heavier tails, meaning it experiences larger daily swings in both directions. This greater volatility helps explain why behavioral signals relate more strongly to Ethereum's movements."
#| message: false
#| warning: false

ggplot(df, aes(x = daily_return, fill = coin)) +
  geom_density(alpha = 0.6, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_continuous(labels = label_percent(), limits = c(-0.15, 0.15)) +
  scale_fill_manual(values = c("BITCOIN" = "#3182bd", "ETHEREUM" = "#e6550d")) +
  labs(
    x = "Daily Return",
    y = "Density",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

Ethereum's distribution is wider than Bitcoin's with heavier tails on both sides. This confirms what the earlier analysis suggested: Ethereum experiences larger daily swings in both directions. Its sharper sentiment swings, higher volatility, and stronger response to lagged behavioral signals all trace back to this fundamental difference. Bitcoin's tighter distribution reflects a more stable market that is less sensitive to shifts in online mood and attention.

------------------------------------------------------------------------

## Summary

This exploratory analysis examined how sentiment and search interest relate to Bitcoin and Ethereum price movements. Three questions guided the investigation:

**How do prices, sentiment, and public attention change over time?** All three variables move in recognizable cycles. Prices decline early in the year, rally in summer, and pull back in fall. Sentiment and search interest follow these price cycles, rising during rallies and falling during declines.

**How closely do sentiment and search interest track market volatility?** Sentiment becomes more dispersed on high-volatility days, with both strongly positive and strongly negative readings appearing when price moves are large. Search interest also rises during volatile periods but does not show the same dispersion pattern. Sentiment appears more connected to the emotional character of market activity.

**Do Bitcoin and Ethereum respond differently to behavioral signals?** Yes. Ethereum consistently shows stronger reactions. It has wider sentiment swings, higher volatility, and a more dispersed return distribution. Previous-day sentiment shows a modest positive relationship with next-day returns for Ethereum but almost none for Bitcoin. Search interest provides weak signals for both coins. These differences suggest that Ethereum's market is more sensitive to shifts in online attention and attitudes than Bitcoin's is.
